We are not the 99 percent: the Local Group as an outlier in galaxy pairs' satellite distributions

### 2017-10-27 paper/paper.tex
updated captions + results from the gaussian model

\begin{abstract}
We quantify the joint spatial distribution of the brightest satellites
around the Milky Way (MW) and M31, the dominant galaxies in the Local
Group (LG).
We use the Illustris-1 and ELVIS simulations to show that the
MW satellite distribution is significantly thinner and more oblate
than the expectations from the Lambda Cold Dark Matter (LCDM)
paradigm. In contrast, the satellites in M31 are fully consistent with the
average expectations from simulations.
We also find that the satellite planes in the MW and M31 are almost perpendicular to the line
connecting the two galaxies. While none of the pairs in the simulations show
a similar degree of alignment, these alignments are consistent with a
random uniform distribution, making the observations
consistent with theoretical expectations.
The special MW satellite distribution makes the LG an atypical pair in
the cosmological context provided by LCDM simulations. 
We estimate that only $0.5\%$ to $1\%$ of the pairs with isolation
and dynamical characteristics similar to the LG are expected to have
satellite distributions with the same degree of atipicality, making
the LG a $3\sigma$ outlier in a sample of simulated galaxy pairs.
\end{abstract}

@@

\section{Observational Data}
\label{sec:obs}

\section{Data from the ELVIS project}
\label{sim:ELVIS}

\section{Data from the Illustris project}
\label{sec:illustris}

@@

Although Illustris-1 has stellar particles, we do not use their
properties to select the satelite population because the smallest
galaxies are barely resolved in stellar mass at magnitudes of
$M_V=9$. We prefer using the dark matter information as the smallest
sub-halos are sampled with at least $35$ particles. 
For this reason we select the satellite galaxy samples from the

We chose the satellite samples by ranking the subhalos in decreasing
order of its maximum circular velocity and select the first $N_p$
halos in the list. 
The results in the main body of the paper correspond to $11\leq
N_p\leq 15$.

@@

We also use two kinds of satellite distributions. 
The first keeps the positions for the satellites fixed as provided in
the observations/simulations; the second randomizes the angular positions of
the satellites around the central galaxy while keeping its radial
distance fixed. The randomization process is done 100 times for each
galaxy.

@@

\begin{figure*}
\centering
\includegraphics[width=0.32\textwidth]{scatter_random_ranked_width.pdf}
\includegraphics[width=0.32\textwidth]{scatter_ranked_width.pdf}
\includegraphics[width=0.32\textwidth]{scatter_norm_ranked_width.pdf}
\caption{Plane width characterization in the Local Group and the
  Isolated Pairs. In all panels the horizontal axis corresponds to the
  M31 or the most massive halo in the pair and the vertical axis to
  the MW or the least massive halo in the pair.
The panel on the left shows the plane width in physical units
comparing the results of five measurements from the observations
(stars) and the result of spherically randomizing the satellite
positions (circles). 
The panel in the middle compares the average from the observations
(star) and the average from each one of the Isolated Pairs (circles
with error bars).
The panel on the right has the same information as the middle panel,
only that this time each point has been normalized (median substracted
and normalized by the standar deviation) to the results of its
randomization. 
The main message of this series of plots is that the MW has a
significantly thiner plane both compared to the result of its own
satellite spherical randomization (left panel) and the expectation from
simulations (middle panel). 
This low value is $2\sigma$ away from what is expected in a spherical
distribution. 
In the M31 their satellites are in agreement with the expectations
both from an spherical distribution and the results form the
simulations. 
A second conclusion is that the spherically averaged plane width
MW (seen in the point cloud in the left panel) is smaller than the
average expectation from simulations, while for M31 the spherical
average is consistent with simulations. 
\label{fig:scatter_width}}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=0.32\textwidth]{scatter_random_ranked_mu.pdf}
\includegraphics[width=0.32\textwidth]{scatter_ranked_mu.pdf}
\includegraphics[width=0.32\textwidth]{scatter_norm_ranked_mu.pdf}
\caption{Same layout as in Figure \ref{fig:scatter_width}. 
This time for $\mu$ the absolute value of the dot product between the
vector connecting the main galaxies and the vector perpendicular to
the satellite plane.
In this case both the MW and M31 show a strong alignment.
Apparently this result is outside the expectations from the
simulations and atypical compared to the spherical results. 
However, this is not the case. The $\mu$ distributions in the
randomization and the simulation are consistent with a uniform
distribution.
Under such circumstances $\mu\approx 1$ is as likely as
any other value.
\label{fig:scatter_mu}}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{gaussian_model_illustris_M31.pdf}
\includegraphics[width=0.45\textwidth]{gaussian_model_illustris_MW.pdf}
\caption{Correlations between the multivariate gaussian model built on
  the normalized values for the plane width $w$, $c/a$ ratio and $b/a$ ratio. 
Left/right panel correspond to M31/MW. 
The contour levels in the 2D histograms correspond to the $1\sigma$,
$2\sigma$ and $3\sigma$ contours in two dimensions. 
The dahsed vertical lines in the histograms along the diagonal
correspond to the $1\sigma$ boundaries in one dimension.
The results for the gaussian model are built from $10^6$ point
realizations in the six-dimensional space spaned by the variables of
interest. 
The correlation matrix and the mean values are computed from the
results in the Illustris-1 simulation.
The cross indicates the LG values.
This plot clearly shows how the M31 results are well within the
expectations from simulations while MW has an unusual low value for
the plane width and the $c/a$ axis ration.
Similar results from the ELVIS data are presented in Figure
\ref{fig:correlations_elvis}.
\label{fig:correlations_illustris}}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=0.47\textwidth]{expected_numbers.pdf}
\caption{Probability distribution for the expected number of pairs
  showing the same degree of atipicality as the Local Group. The two
  distributions correspond to different numbers of initial isolated
  pairs. On average $1\%$ of the isolated pairs should present
  satellite distributions as atypical as the Local Group.
\label{fig:expected_number}}
\end{figure}

@@

\section{Discussion}
\subsection{Gaussian Model}

@@

\section{Results from ELVIS}
\begin{figure*}
\centering
\includegraphics[width=0.24\textwidth]{scatter_ranked_elvis_width.pdf}
\includegraphics[width=0.24\textwidth]{scatter_ranked_elvis_ca_ratio.pdf}
\includegraphics[width=0.24\textwidth]{scatter_ranked_elvis_ba_ratio.pdf}
\includegraphics[width=0.24\textwidth]{scatter_ranked_elvis_mu.pdf}
\includegraphics[width=0.24\textwidth]{scatter_norm_ranked_elvis_width.pdf}
\includegraphics[width=0.24\textwidth]{scatter_norm_ranked_elvis_ca_ratio.pdf}
\includegraphics[width=0.24\textwidth]{scatter_norm_ranked_elvis_ba_ratio.pdf}
\includegraphics[width=0.24\textwidth]{scatter_norm_ranked_elvis_mu.pdf}
\caption{Same layout as in Figure \ref{fig:scatter_width}. 
This time for the $b/a$ axis ratio. In this case both the MW and M31
are consistent with the results of a spherical distribution and the
simulations. 
\label{fig:scatter_ba_ratio}}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{gaussian_model_elvis_M31.pdf}
\includegraphics[width=0.45\textwidth]{gaussian_model_elvis_MW.pdf}
\caption{Correlations between the normalized values for the plane
  width, $c/a$ ratio and $b/a$ ratio. 
The star corresponds to the LG, black circles with errorbars are the
results from simulations and the gray cloud is the resul from the
multivariate gaussian model.
Upper/lower row summarizes the results for M31/M31.
This simplified description allows us to quantify how atypical is the
LG compared to the simulation results. 
\label{fig:correlations_elvis}}
\end{figure*}

>>>>

\begin{abstract}
In this paper we quantify the joint spatial distribution of the brightest satellites
around the Milky Way (MW) and M31, the dominant galaxies in the Local
Group (LG).
We use the Illustris-1 and ELVIS simulations to quantify how the 
unusual satellite distribution in the MW  makes the LG an atypical pair in
the broader cosmological context provided by LCDM simulations. 
We find that only $0.05\%$ to $0.25\%$ of the pairs with isolation
and dynamical characteristics similar to the LG are expected to have
satellite distributions with the same degree of atipicality, making
the LG a $3\sigma$ outlier.
\end{abstract}

@@

\label{Data Samples}

\subsection{Observational Data}
\label{sec:obs}

\subsection{Data from the ELVIS project}
\label{sim:ELVIS}

\subsection{Data from the Illustris project}
\label{sec:illustris}

@@

Although Illustris-1 has stellar particles, we do not use their
properties to select the satelite population because the smallest
galaxies are barely resolved in stellar mass at magnitudes of
$M_V=9$. We prefer using the dark matter information as the smallest
sub-halos are sampled with at least $35$ particles. 
We chose the satellite samples by ranking the subhalos in decreasing
order of its maximum circular velocity and select the first $N_p$
halos in the list. 
The results presented here correspond to $11\leq N_p\leq 15$.

@@

We also use two kinds of satellite distributions. 
The first keeps the positions for the satellites fixed as provided in
the observations/simulations; the second randomizes the angular positions of
the satellites around the central galaxy while keeping its radial
distance fixed. The randomization process is done 1000 times for each
galaxy.

@@

\begin{figure*}
\centering
\includegraphics[width=0.32\textwidth]{scatter_random_ranked_width.pdf}
\includegraphics[width=0.32\textwidth]{scatter_ranked_width.pdf}
\includegraphics[width=0.32\textwidth]{scatter_norm_ranked_width.pdf}
\caption{Plane width characterization in the Local Group and the
  Isolated Pairs. In all panels the horizontal axis corresponds to the
  M31 or the most massive halo in the pair and the vertical axis to
  the MW or the least massive halo in the pair.
The panel on the left shows the plane width in physical units
comparing the results from observations
(stars) against the result of spherically randomizing the satellite
positions while keeping its radial distance (circles). 
The panel in the middle compares the average from the observations
(star) and the average from each one of the Isolated Pairs (circles
with error bars).
The panel on the right has the same information as the middle panel,
only that this time each point has been normalized (median substracted
and normalized by the standar deviation) to the results of its
randomization. 
The main message of this series of plots is that the MW has a
significantly thiner plane both compared to the result of its own
satellite spherical randomization (left panel) and the expectation from
simulations (middle panel). 
This low value is $2\sigma$ away from what is expected in a spherical
distribution. 
In the M31 their satellites are in agreement with the expectations
both from an spherical distribution and the results form the
simulations. 
A second conclusion is that the spherically averaged plane width
MW (seen in the point cloud in the left panel) is smaller than the
average expectation from simulations, while for M31 the spherical
average is consistent with simulations. 
\label{fig:scatter_width}}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=0.32\textwidth]{scatter_random_ranked_mu.pdf}
\includegraphics[width=0.32\textwidth]{scatter_ranked_mu.pdf}
\includegraphics[width=0.32\textwidth]{scatter_norm_ranked_mu.pdf}
\caption{Same layout as in Figure \ref{fig:scatter_width}. 
This time for $\mu$ the absolute value of the dot product between the
vector connecting the main galaxies and the vector perpendicular to
the satellite plane.
In this case both the MW and M31 show a strong alignment.
Apparently the LG is outside the expectations from
simulations and atypical compared to the spherical results. 
However, this is not the case. The $\mu$ distributions in the
randomization and the simulation are consistent with a uniform
distribution.
Under such circumstances $\mu\approx 1$ is as likely as
any other value.
\label{fig:scatter_mu}}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{gaussian_model_illustris_M31.pdf}
\includegraphics[width=0.45\textwidth]{gaussian_model_illustris_MW.pdf}
\caption{Correlations between the multivariate gaussian model built on
  the normalized values for the plane width $w$, $c/a$ ratio and $b/a$ ratio. 
Left/right panel correspond to M31/MW. 
The contour levels in the 2D histograms correspond to the $1\sigma$,
$2\sigma$ and $3\sigma$ contours in two dimensions. 
The dahsed vertical lines in the histograms along the diagonal
correspond to the $1\sigma$ boundaries in one dimension.
The results for the gaussian model are built from $10^6$ point
realizations in the six-dimensional space spaned by the variables of
interest. 
The correlation matrix and the mean values are computed from the
results in the Illustris-1 simulation.
The cross indicates the LG values.
This plot clearly shows how the M31 results are well within the
expectations from simulations while MW has an unusual low value for
the plane width and the $c/a$ axis ratio.
Equivalent results from the ELVIS data are presented in Figure
\ref{fig:correlations_elvis}.
\label{fig:correlations_illustris}}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=0.47\textwidth]{expected_numbers.pdf}
\caption{Probability distribution for the expected number of pairs
  showing the same degree of atipicality as the Local Group if drawn
  from a sample of 2000 isolated pairs. 
  The distributions correspond to results derived from Illustris-1
  and ELVIS data.
  On average between $0.05\%$ and $0.25\%$ of the isolated pairs should present
  satellite distributions as atypical as the Local Group.
\label{fig:expected_number}}
\end{figure}

@@

\section{Discussion}
\subsection{Outlier characterization with a Multivariate Gaussian Model}

@@

\section{Physical Characteristics of the Isolated Pairs samples}

\section{Results from ELVIS}
\begin{figure*}
\centering
\includegraphics[width=0.24\textwidth]{scatter_ranked_elvis_width.pdf}
\includegraphics[width=0.24\textwidth]{scatter_ranked_elvis_ca_ratio.pdf}
\includegraphics[width=0.24\textwidth]{scatter_ranked_elvis_ba_ratio.pdf}
\includegraphics[width=0.24\textwidth]{scatter_ranked_elvis_mu.pdf}
\includegraphics[width=0.24\textwidth]{scatter_norm_ranked_elvis_width.pdf}
\includegraphics[width=0.24\textwidth]{scatter_norm_ranked_elvis_ca_ratio.pdf}
\includegraphics[width=0.24\textwidth]{scatter_norm_ranked_elvis_ba_ratio.pdf}
\includegraphics[width=0.24\textwidth]{scatter_norm_ranked_elvis_mu.pdf}
\caption{ELVIs results for the quantities presented for the Illustris-1
  simulation in Figures  \ref{fig:scatter_width},
  \ref{fig:scatter_ca_ratio}, \ref{fig:scatter_ba_ratio} and
  \ref{fig:scatter_mu}.
Upper row corresponds to the raw values from observations and
simulated pairs, while the second row normalizes the same values to
the mean and standard deviation on its spherically randomized
counterparts. 
\label{fig:scatter_elvis}}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{gaussian_model_elvis_M31.pdf}
\includegraphics[width=0.45\textwidth]{gaussian_model_elvis_MW.pdf}
\caption{
Same layout as Figure \ref{fig:correlations_illustris}, this time
computed from the ELVIS data.
\label{fig:correlations_elvis}}
\end{figure*}
